---
title: "SSF-Exercise-2"
output: html_document
---

Using the Otters_SSF.html and OttersTwo_step.html as resources:

1. Fit a mixed SSF model to the fisher data.
2. Fit models to individuals and estimate typical (i.e. mean) coefficients for 
`popden`, `forest`, and `elevation`, along with a confidence interval for the mean.

To help, I've provided code that you can use to read in and process the data below, and you can look at a solution (`MultipleAnimalsSSFs.R`) in the output folder.

# Setup

Load libraries

```{r warning=FALSE, message=FALSE}
library(glmmTMB)
library(tidyverse) 
library(survival)
library(TwoStepCLogit)
library(amt)
library(terra)
library(here)
library(broom) 
library(tictoc)  
options(width=165)
knitr::opts_chunk$set(fig.width=12,fig.height=4.5, error=TRUE)
set.seed(09081940)
```

Record time for running all code

```{r}
tic("total")
```

## Environmental data

Read in and process the environmental data layers (elevation, population density. forest)

```{r}
# Read in data
elevation <- rast(here("Resource-and-Step-Selection-Analyses/Data/elevation/", "ASTER ASTGTM2 Elevation-20100101120000000-0-0.tif"))
landuse <- rast(here("Resource-and-Step-Selection-Analyses/Data/landuse", "landuse_study_area.tif"))
popden <- rast(here("Resource-and-Step-Selection-Analyses/Data/pop_den", "pop_den.tif"))

# Center and scale elevation and popden
elevation[] <- (elevation[] - mean(elevation[], na.rm = TRUE))/sd(elevation[], na.rm = TRUE)
popden[] <- (popden[] - mean(popden[], na.rm = TRUE))/sd(popden[], na.rm = TRUE)

# Reproject rasters to EPSG:5070
landuse <- project(landuse, "EPSG:5070")
elevation <- project(elevation, landuse)
popden <- project(popden, landuse)

# Create a binary layer where landuse classes 41 to 43 represent forest
forest <- landuse %in% 41:43
names(elevation)<-"elevation"
names(forest)<-"forest"
names(popden)<-"popden"
```

Plot the layers

```{r}
plot(landuse)
plot(elevation)
plot(popden)
```

## Fisher data

Load the fisher data  

```{r eval=TRUE}
dat <- read_csv(here("Resource-and-Step-Selection-Analyses/Data", "fisher_data.csv")) %>% 
  filter(id %in% c(1465, 1466, 1072, 1078, 1016, 1469)) 
```

We will make use of nested data frames in R, which will allow us to apply functions
separately to data from each individual using the map functions in the Purr pacakge.
If you are unfamiliar with `purrr` syntax, may want to view one or more of the tutorials, 
below, or make use of the [purrr cheat sheet](https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf).

- http://www.rebeccabarter.com/blog/2019-08-19_purrr/
- https://www.r-bloggers.com/2020/05/one-stop-tutorial-on-purrr-package-in-r/
- https://jennybc.github.io/purrr-tutorial/index.html

```{r}
dat_all <- dat %>% nest(data=c(x,y,t)) 
```

Add sex as a column of the nested data frame

```{r}
dat_all$sex <- c("f", "f", "f", "m", "m", "m")
```

Create tracks with an appropriate coordinate reference system
using the amt package

```{r}
dat_all <- dat_all %>% 
  mutate(trk = map(data, function(d) {
    make_track(d, x, y, t, crs = 4326) %>% 
      transform_coords(crs_to = 5070)
  }))
dat_all
```

Summarize sampling rates, 

```{r}
dat_all %>% mutate(sr = map(trk, summarize_sampling_rate)) %>% 
  dplyr::select(id, sr) %>% unnest(cols=c(sr))
```

10 minutes seems to appropriate for all animals.

# Prepare data for fitting step-selection functions

First, we will create a semi-regular trajectory by subsampling the track 
to include locations that are separated by 10 +/- 2 minutes.

```{r}
dat1 <- dat_all %>% mutate(dat_clean = map(trk, ~ {
  .x %>% track_resample(rate = minutes(10), tolerance = seconds(120))
}))
```

Rather than fit tentative step length and turn angle distributions to 
each individual separately, we will pool the data and fit a single gamma
distribution to the **pooled** step length data and a single von-Mises
distribution to the pooled turn angle data.  

We will then generate random steps using these distributions and include
step length, log(step length), and cos(turn angle) in the model as 
both fixed and random effects.  That will allow each individual to have its
own movement kernel with the fixed effects used to update the movement kernel
for a "typical" individual (one with all random effects set to 0).  The random
effects will  be used to update the movement kernel for all observed
individuals in the data set. These updates can be accomplished using the 
[mixedSSA package](https://github.com/smthfrmn/mixedSSA). Here, we will largely
focus on the habitat selection parameters and the mean correction factors
for the movement parameters (needed to estimate the movement kernel for a 
"typical" individual).


First, create a steps version of the data set using the steps_by_burst 
function applied to the data from each individual. We will unnest the
data so that we can fit a single set of distributions to the pooled step
length and turn angle data. 

```{r}
dat_ssf <- dat1 %>% 
  mutate(stps = map(dat_clean, ~ .x %>% 
                      steps_by_burst())) %>%
  select(id, sex, stps) %>%
  unnest(stps)
```

Unfortunately, some of the classes were lost in the process, so we have to 
add some of the classes back on to the object.

```{r}
class(dat_ssf) <- c("bursted_steps_xyt", "steps_xyt", "steps_xy", class(dat_ssf)) 
```

Now, we can pool the steps across individuals to fit a single tentative 
distribution to the step lengths and turn angles, and use this distribution
to generate random steps.  Also, we will extract the covariates at the 
end of the observed and random steps.

```{r}
dat_ssf2 <- dat_ssf %>% random_steps() %>%
  extract_covariates(elevation) %>%
  extract_covariates(popden) %>%
  extract_covariates(forest) %>%
  mutate(
    y = as.numeric(case_),
    id = as.numeric(factor(id)), 
    cos_ta_= cos(ta_),
    log_sl_ = log(sl_),
    step_id = paste(id, step_id_, sep = ":")) %>%
    dplyr::select(id, sex, y, step_id, elevation, popden, forest,
                  cos_ta_, log_sl_, sl_, ta_)
```

create forest indicator variable (1 if forest, 0 otherwise)

```{r}
dat_ssf2$forest <- as.numeric(dat_ssf2$forest)
```

At this point, you should have everything you need for fitting a mixed SSA using 
glmmTMB (or for fitting models to individuals).


