---
title: "Step-selection function: Exericise 1"
output: html_document
---

For this exercise, you will work with Scott Lapoint's fisher data:

LaPoint, S, Gallery P, Wikelski M, Kays R (2013) Animal behavior, cost-based corridor models, and real corridors. Landscape Ecology, v 28 i 8, p 1615–1630. doi:10.1007/s10980-013-9910-0

LaPoint S, Gallery P, Wikelski M, Kays R (2013) Data from: Animal behavior, cost-based corridor models, and real corridors. Movebank Data Repository. doi:10.5441/001/1.2tp2j43g

Or, alternatively, feel free to work with your own data!

Let's load packages we will need.

```{r warning = FALSE, message = FALSE}
library(amt)
library(terra)
librayr(tidyverse)
library(here)
```

It will help to have the `amt_demo_iSSF.html` file open as you work through the 
exercise.

First, we will read in the data and environmental layers capturing elevation, 
population density, and landuse categories.  We will center and scale the 
covariate data and also create an indicator variable that is 1 if the 
observation is in forest and 0 otherwise.

```{r}
# Location data 
dat <- read_csv(here("Resource-and-Step-Selection-Analyses/Data", "fisher_data.csv")) 

# Environmental layers
elevation <- rast(here("Resource-and-Step-Selection-Analyses/Data/elevation/", "ASTER ASTGTM2 Elevation-20100101120000000-0-0.tif"))
landuse <- rast(here("Resource-and-Step-Selection-Analyses/Data/landuse", "landuse_study_area.tif"))
popden <- rast(here("Resource-and-Step-Selection-Analyses/Data/pop_den", "pop_den.tif"))

# Center and scale elevation and popden
elevation[] <- (elevation[] - mean(elevation[], na.rm = TRUE))/sd(elevation[], na.rm = TRUE)
popden[] <- (popden[] - mean(popden[], na.rm = TRUE))/sd(popden[], na.rm = TRUE)
 
# Reproject rasters to EPSG:5070
landuse <- project(landuse, "EPSG:5070")
elevation <- project(elevation, "EPSG:5070")
popden <- project(popden, "EPSG:5070")

# Create a binary layer where landuse classes 41 to 43 represent forest
forest <- landuse %in% 41:43
plot(forest)

# Rename the layers
names(elevation)<-"elevation"
names(forest)<-"forest"
names(popden)<-"popden"
```



1. Filter the data to include observations from one of the individuals (feel free to choose a different individual).
```{r}
dat <- dat %>% 
  filter(id == 1078) 
dat
```

2. Create a track using the `make_track` function.
```{r}

```

3. Check the sampling rate of the observations using the `summarize_sampling_rate`.
```{r}

```


4. Use `track_resample` to choose an appropriate sampling frequency to model movements.
```{r}

```

5. Create a *steps* version fo the data using the `steps_by_burst` function.
```{r}

```


5. Plot the empirical distribution of step lengths and turn angles. 

```{r}

```


6. Use the `random_steps` function to:  a) fit a gamma distribution to the step 
lenghts and a von-Mises distribution to the turn angles, and b) generate 
20 random steps for each observed step. 

```{r}

```


7. Extract covariates for population, elevation, and forest.  You can choose
whether to extract these covariate values at the start of the step, end of the 
step, or both.
```{r}

```

8. Fit a step-selection model to the data form step 7.
```{r}

```

9. Interpret the parameters in terms of relative probability of use. Hint: look 
at a summary of the model using the `summary` function.

```{r}

```

10. Plot updated step length and turn angle distributions.

```{r}

```

