---
title: "Step-selection function: Exericise 1"
output: html_document
---

For this exercise, you will work with Scott Lapoint's fisher data:

LaPoint, S, Gallery P, Wikelski M, Kays R (2013) Animal behavior, cost-based corridor models, and real corridors. Landscape Ecology, v 28 i 8, p 1615–1630. doi:10.1007/s10980-013-9910-0

LaPoint S, Gallery P, Wikelski M, Kays R (2013) Data from: Animal behavior, cost-based corridor models, and real corridors. Movebank Data Repository. doi:10.5441/001/1.2tp2j43g

Or, alternatively, feel free to work with your own data!

Let's load packages we will need.

```{r warning = FALSE, message = FALSE}
library(amt)
library(terra)
library(tidyverse)
library(here)
```

It will help to have the `amt_demo_iSSF.html` file open as you work through the 
exercise.

First, we will read in the data and environmental layers capturing elevation, 
population density, and landuse categories.  We will center and scale the 
covariate data and also create an indicator variable that is 1 if the 
observation is in forest and 0 otherwise.

```{r}
# Location data 
dat <- read_csv(here("Resource-and-Step-Selection-Analyses/Data", "fisher_data.csv")) 

# Environmental layers
elevation <- rast(here("Resource-and-Step-Selection-Analyses/Data/elevation/", "ASTER ASTGTM2 Elevation-20100101120000000-0-0.tif"))
landuse <- rast(here("Resource-and-Step-Selection-Analyses/Data/landuse", "landuse_study_area.tif"))
popden <- rast(here("Resource-and-Step-Selection-Analyses/Data/pop_den", "pop_den.tif"))

# Center and scale elevation and popden
elevation[] <- (elevation[] - mean(elevation[], na.rm = TRUE))/sd(elevation[], na.rm = TRUE)
popden[] <- (popden[] - mean(popden[], na.rm = TRUE))/sd(popden[], na.rm = TRUE)
 
# Reproject rasters to EPSG:5070
landuse <- project(landuse, "EPSG:5070")
elevation <- project(elevation, "EPSG:5070")
popden <- project(popden, "EPSG:5070")

# Create a binary layer where landuse classes 41 to 43 represent forest
forest <- landuse %in% 41:43

# Plot forest
plot(forest)

# Rename the layers
names(elevation)<-"elevation"
names(forest)<-"forest"
names(popden)<-"popden"

```



1. Filter the data to include observations from one of the individuals (feel free to choose a different individual).
```{r}
dat <- dat %>% 
  filter(id == 1072) 
dat
```

2. Create a track using the `make_track` function. Use the arguement crs = 4326. 
Then, use transform_coords to transform the track to crs 5070.
```{r}
trk <- make_track(dat, .x = x, .y = y, .t = t, crs = 4326)
trk <- trk %>% transform_coords(crs_to = 5070)
```

3. Check the sampling rate of the observations using the `summarize_sampling_rate`.
```{r}
summarize_sampling_rate(trk)
```


4. Use `track_resample` to choose an appropriate sampling frequency to model movements.
Note: there are many options here, but the median time interval is 10 minutes (so, that is the shortest sampling frequency we should choose).
```{r}
trk2 <-track_resample(trk, rate = minutes(10), tolerance = minutes(1))
```

5. Create a *steps* version fo the data using the `steps_by_burst` function.
```{r}
fishersteps <- steps_by_burst(trk2)
```


5. Plot the empirical distribution of step lengths and turn angles. 

The distribution of step lengths is highly right skewed. There are some really large
step lengths that should probably be looked at to make sure they are reasonable
(these may be due to large measurement errors).
```{r}
ggplot(fishersteps, aes(x = sl_)) + 
  geom_histogram(aes(y = ..density..), binwidth = 2)  
```

The distribution of turn angles is multi-modal with modes at 0 and +/- $\pi$ (as if often the case).
```{r}
ggplot(fishersteps, aes(x = ta_)) + 
  geom_histogram(aes(y = ..density..), binwidth = 0.2)  
```


6. Use the `random_steps` function to:  a) fit a gamma distribution to the step 
lenghts and a von-Mises distribution to the turn angles, and b) generate 
20 random steps for each observed step. 

```{r}
obsavail <- random_steps(fishersteps, n_control = 20)
```


7. Extract covariates for population, elevation, and forest.  You can choose
whether to extract these covariate values at the start of the step, end of the 
step, or both.
```{r}
ssfdat <- obsavail %>%
  extract_covariates(elevation, where = "both") %>%
  extract_covariates(forest, where = "end") %>%
  extract_covariates(popden, where = "end") %>%
  mutate(log_sl_ = log(sl_))
```

8. Fit a step-selection model to the data form step 7.
```{r}
m1 <- fit_issf(ssfdat, 
               case_ ~ elevation_end + forest + popden + 
                 sl_ + log_sl_ + cos(ta_) + (sl_+log_sl_):elevation_start +
               strata(step_id_),
               model=TRUE)
summary(m1)
```

9. Interpret the parameters in terms of relative probability of use.

Broadly: the fisher appears to be selecting for higher elevations 
(positive coefficient, but not statistically significant),
forest (positive coefficient), and lower population densities (negative coefficient).  
The step lengths also depend on elevation at the starting location.

Looking at the exponentiated coefficient for forest, we can conclude that
if presented with two identical observations (with respect to elevation,
population density, and distance from the fisher), the fisher would be 2.29 times
more likely to choose the location in the forest. If we exponentiate the coefficients
for elevation and popden, we can state similar conclusions (but they represent
comparisons that differ by 1 in these variables).  Since the variables were 
centered and scaled, we would conclude the fisher would be 0.57 times less likely to
choose a location that was 1 sd higher in population density, but otherwise had 
the same elevation and landuse type. 
 

10. Plot updated step length and turn angle distributions.

Tentative distribution
```{r}
tent_sl <- sl_distr(m1)
```

Because we have an interaction between the step length variables and a continuous variable, 
we have to use the `update_gamma` function as outlined in [Appendix B of Fieberg et al. 2021](https://conservancy.umn.edu/server/api/core/bitstreams/63727072-87b1-4b35-b81c-8fd31b8f1e57/content).

Updated distribution when elevation_start = 25th percentile 

```{r}
elevs.025 <- quantile(ssfdat$elevation_start, prob = 0.025)
upd_sl_low <- update_gamma(dist = m1$sl_,  
                           beta_sl = m1$model$coefficients["sl_"] +
                               elevs.025*m1$model$coefficients["sl_:elevation_start"],
                           beta_log_sl = m1$model$coefficients["log_sl_"] +
                               elevs.025*m1$model$coefficients["log_sl_:elevation_start"]) 
```
Now, get the distribution at the 75% of elevation_start

```{r}
elevs.975 <- quantile(ssfdat$elevation_start, prob = 0.975)
upd_sl_high <- update_gamma(dist = m1$sl_,  
                           beta_sl = m1$model$coefficients["sl_"] +                       
                              elevs.975*m1$model$coefficients["sl_:elevation_start"],
                           beta_log_sl = m1$model$coefficients["log_sl_"] +     
                             elevs.975*m1$model$coefficients["log_sl_:elevation_start"]) 
```
Put these all together and plot the distributions.

```{r}
tent_df <- data.frame(dist = "tent",
                      shp = tent_sl$params$shape,
                      scl = tent_sl$params$scale)
upd_high <- data.frame(dist = "upd High",
                     shp = upd_sl_high$params$shape,
                     scl = upd_sl_high$params$scale)
upd_low <- data.frame(dist = "upd Low",
                     shp = upd_sl_low$params$shape,
                     scl = upd_sl_low$params$scale)
(sl_df <- rbind(tent_df, upd_high, upd_low))

expand.grid(sl = seq(1, 1500, length.out = 100),
            dist = c("tent", "upd High", "upd Low")) %>% 
  left_join(sl_df) %>% 
  mutate(y = dgamma(sl, shape = shp, scale = scl)) %>% 
  ggplot(aes(x = sl, y = y, color = dist)) +
  geom_line() +
  xlab("Step Length (m)") +
  ylab("Probability Density") +
  theme_bw()
```

Mean step length for the gamma = shape * scale parameter
```{r}
sl_df$mean = sl_df$shp*sl_df$scl
sl_df
```

Now, let's look at the turn angle distribution.  First, we update the parameters.
```{r}
tent_ta <- ta_distr(m1)
upd_ta <- update_ta_distr(m1, beta_cos_ta = "cos(ta_)")
tent_df_ta <- data.frame(dist = "tent",
                         mu = tent_ta$params$mu,
                         k = tent_ta$params$kappa)
(upd_df_ta <- data.frame(dist = "upd",
                        mu = upd_ta$params$mu,
                        k = upd_ta$params$kappa))


```

The von Mises is not defined for negative kappa. Negative kappa indicates that the turns are not actually concentrated around 0 (as we typically assume), but rather are concentrated around +/- pi. If this happens, we can simply multiply the estimated kappa by -1 and change mu to pi.

```{r}
upd_df_ta <- data.frame(dist = "upd",
                        mu = pi,
                        k = -1 * upd_ta$params$kappa)
(ta_df <- rbind(tent_df_ta, upd_df_ta))
```

Plot the distribution.

```{r}
expand.grid(ta = seq(-pi, pi, length.out = 100),
            dist = c("tent", "upd")) %>% 
  left_join(ta_df) %>% 
  # circular::dvonmises is not vectorized
  rowwise() %>% 
  mutate(y = circular::dvonmises(ta, mu = mu, kappa = k)) %>% 
  ggplot(aes(x = ta, y = y, color = dist)) +
  geom_line() +
  xlab("Turn Angle (radians)") +
  ylab("Probability Density") +
  scale_x_continuous(breaks = c(-pi, -pi/2, 0, pi/2, pi),
                     labels = expression(-pi, -pi/2, 0, pi/2, pi)) +
  coord_cartesian(ylim = c(0, 0.4)) +
  theme_bw()
```

